---
title: "Meta profile analysis of Rrm4 crosslink signal across stop codon"
date: last-modified
author: 
    - name:  "Srimeenakshi Sankaranarayanan"
      
title-block-banner: true
format: 
    html:
        theme: flatly
        self-contained: true
        code-fold: true
        code-summary: "Show the code"
        toc: true
        number-sections: true
        anchor-sections: true
    pdf:
        toc: true
        number-sections: true
        self-contained: true
editor: source
execute:
  echo: true
  error: false
  warning: false
  message: false
---
Previosly, Rrm4 was shown to bind specifically at the stop codon, suggesting a link between RNA transport and translation. Here, I will assess whether the improved iCLIP2 protocol detects stop-codon binding with higher efficiency than the previous iCLIP protocol used in Olgeiser et.al (2019). 

```{r}
# Libraries
library(BSgenome.Umaydis.ENSMBL.UM1) # forged tomato genome
library(ggplot2)
library(BindingSiteFinder)
library(rtracklayer)
library(ComplexHeatmap)
library(GenomicFeatures)
library(forcats)
library(tidyr)
library(dplyr)
library(tidyverse)
library(GenomicRanges)
library(magick)
library(magrittr)
library(gridExtra)
library(IRanges)
library(Biostrings)
library(ggrepel)
library(ggpubr)
library(viridis)
library(reshape2)
library(gprofiler2)
library(ggsci)
```

```{r}
## theme for ggplot
my_theme = theme_set(theme_bw() +
          theme(axis.text = element_text(size = 10, colour="black"),
                axis.title = element_text(size=12, colour="black"),
                axis.ticks=element_line(color="black"),
                axis.ticks.length=unit(.15, "cm"),
                panel.border=element_rect(color="black", fill = NA),
                panel.background = element_blank(),
                plot.background = element_blank(),
                legend.text = element_text(size=10),
                legend.position = "bottom",
                aspect.ratio = 1/1)
)
```

```{r}
## load gene annotation
gnsN = readRDS("~/Documents/Bioinfo_analysis/iCLIP_course/diffRNA_binding/Annotation_preparation/gns.rds")

trl = readRDS("~/Documents/Bioinfo_analysis/iCLIP_course/diffRNA_binding/Annotation_preparation/trl.rds")
```


```{r}
#| warning: false
#| message: false
## Load binding sites
ic2_bs = import("~/Documents/Bioinfo_analysis/iCLIP_course/iCLIP2_Methods_Stoffel_etal_2023/Rrm4_BS_9nt.bed", format="bed")
ic2_bs = resize(ic2_bs, fix="center", width = 1) # resize to center
```


```{r}
#| warning: false
#| message: false

## import crosslink signals from iCLIP1 (olgeiset et.al 2019)
my.path.iclip1 = "~/Documents/Bioinfo_analysis/iCLIP_course/iCLIP2_Methods_Stoffel_etal_2023/iCLIP1/bw"
# plus strand
plus_bw_i1 = list.files(my.path.iclip1, pattern="Rrm4G_plus", full.names = T)
plus_bw_i1 = import(plus_bw_i1, as= "Rle")
# minus strand
minus_bw_i1 = list.files(my.path.iclip1, pattern="Rrm4G_minus", full.names = T)
minus_bw_i1 = import(minus_bw_i1, format="bw")
# make a list
iCLIP1 = list(plus_ic1 = plus_bw_i1,
                minus_ic1 = minus_bw_i1) %>% as("GRangesList")
```

It is important to note that the `Ensembl` gene annotation label chromosomes numerically, whereas the manually curated annotation used in Olgeiser. et al names chromosome as `Um_chrN`, where N represents the chromosome number. To enable comparison between the two iCLIP dataset, I will remove the `Um_chr` prefix from the chromosome name. 

```{r}
# change the chromosome names
iCLIP1_bw_mod = lapply(iCLIP1, function(i){
  # change Um_chr0
  seqlevels(i) = sub("Um_chr0", "", seqlevels(i))
  # change Um_chr
  seqlevels(i) = sub("Um_chr", "", seqlevels(i))
  
  # change Um_chr
  seqlevels(i) = sub("Um_scaf", "um_scaf", seqlevels(i))
  
  return(i)
})

# calculate coverage
iCLIP1_bw_mod = lapply(iCLIP1_bw_mod, function(i){
  cov = coverage(i, weight=i$score)
  return(cov)
})
```

```{r}
#| warning: false
#| message: false

## Import crosslink signals produced from iCLIP2 
my.path.iclip2 = "~/Documents/Bioinfo_analysis/iCLIP_course/datasets/imb_koenig_2021_11_stoffel_iCLIP_ustilago_Rrm4/merged/bw"
# in plus strand
plus_bw_i2 = list.files(my.path.iclip2, pattern="Rrm4G.v3.duprm.plus.bw", full.names = T)
plus_bw_i2 = import(plus_bw_i2, as = "Rle")


# in minus strand
minus_bw_i2 = list.files(my.path.iclip2, pattern="Rrm4G.v3.duprm.minus.bw", full.names = T)
minus_bw_i2 = import(minus_bw_i2, as="Rle")
```


```{r}
## combine the iCLIP data within a list
cl_Rle = list(iC1_p = iCLIP1_bw_mod$plus_ic1,
              iC1_m = iCLIP1_bw_mod$minus_ic1,
              iC2_p = plus_bw_i2,
              iC2_m = minus_bw_i2) 

```

```{r}
#| warning: false
#| message: false
## get the window around stop codon
# get protein coding gene alson
gnsNP = gnsN[gnsN$gene_biotype=="protein_coding",]

# collect the stop ntd position
stops= flank(gnsNP, -303,start = FALSE) %>% flank(.,-3) 

# get the last nucleotide of stop codon
stops = flank(stops,-1, start=FALSE)
# sanity check 2
#getSeq(Umaydis,stops)

# make stop window
stops = stops+30

```

iCLIP signal across the selected window will be normalized using `min-max normalization` as described in Molitor et al. (2022).

```{r}

# normalization of iCLIP signal was carried out as described in Molitor et.al 2022
min_max_noramlize <- function(x){
  x_new <- (x-min(x)) / (max(x)-min(x)) 
  return(x_new)
}
```


```{r}
#| warning: false
#| message: false

## create a metaprofile
# get the crosslinks within the stop window
cl_Win =  c(map(cl_Rle[c(1,3)],~.x[stops[strand(stops)=="+"]] %>% as.matrix()),
            map(cl_Rle[c(2,4)], ~.x[stops[strand(stops)=="-"]] %>% as.matrix()  %>% .[,ncol(.):1]))

cl_win_bin = list(ic1= rbind(cl_Win$iC1_p, cl_Win$iC1_m),
                  ic2=rbind(cl_Win$iC2_p, cl_Win$iC2_m))

# normalize
cl_win_norm = lapply(cl_win_bin, function(i) apply(i,1, min_max_noramlize) %>% t(.))

# sum of the signal at each poition for each experiment
meta_prof_sum = data.frame(ic1= colSums(cl_win_norm$ic1, na.rm=T),
                           ic2 = colSums(cl_win_norm$ic2, na.rm=T),
                           pos = -floor(ncol(cl_win_norm[[1]])/2): floor(ncol(cl_win_norm$ic1)/2))

meta_profiles_sum_gg <- meta_prof_sum %>% reshape2::melt(id.vars = "pos") 

pl = ggplot(meta_profiles_sum_gg, aes(x =pos, y = value, color = variable))+
  geom_line(linewidth=0.5)+
  geom_point(size=1)+
  scale_color_manual(values=c("#A0A0A4", "#00AAAA"), labels = c("iCLIP", "iCLIP2"))+
  labs(x="position relative to stop codon",
       y="normalized iCLIP signal")+
  ylim(0,1500)+
  scale_x_continuous(breaks=seq(-30,30, 10))+
  theme_bw()+
  theme(legend.title = element_blank(), legend.position = "top")

pl

## ggsave
#ggsave("C:/Users/Sri/Documents/Bioinfo_analysis/iCLIP_course/iCLIP2_Methods_Stoffel_etal_2023/metaprofile.pdf", pl,  width = 10, height = 10, units ="cm")
```

Notably, this analysis revealed that iCLIP2 captures Rrm4 crosslinking sites around the stop codon with substantially higher efficiency than the previous iCLIP protocol, suggesting that microbial iCLIP2 improves the mapping of RNA–protein interactions.


```{r}
sessionInfo()
```

